A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-02-18 papers/execution-model/model.tex
sec 3

We write $\emptyset$ for the empty set and $\mathbf{e :: s}$ to
decompose a set into an element $\mathbf{e}$ and a set  $\mathbf{s}$.

@@

The local node offers operations to transfer tokens from one account
to another, to invoke a contract, and to originate a contract on the blockchain.
\begin{align*}
  \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
  \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
\end{align*}
Here, \NTEZ\ and \MTEZ\ are tokens, \PUK\ is the public key of the
sender, \ADDR\ is either a public 
key for an implicit account (in case of a simple transfer)  or a
public hash for a smart contract (for an invocation), \PARAMETER\ is
the argument passed to the smart contract, 
which is empty for a simple transfer. Furthermore,  \CODE\ is
a script of a smart contract and \INIT\ is the initial value of the
contract's storage. Each operation returns an \emph{operation hash}
\OPH, on which we can query the status of the operation.

The local node also offer several ways to query the current state of
the blockchain.
\emph{Queries} are defined by the following grammar:
\begin{align*}
  \QRY & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE\ADDR
  \mid \GETSTATUS\OPH
  \mid \GETSTORAGE\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
  \mid \GETCONTRACT\OPH
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the code of a contract by the
operation hash of its originating transaction.

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR
  \\& \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  \\&\mid \QOP\  \EXPR \mid
  \CONTRACT\ \TYPE\ \EXPR \mid \TOSTRING\
  \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
  &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR\\
  \QOP & ::= balance \mid status \mid storage \mid code \mid counter
         \mid type
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
                  \TOPH \mid
                  \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by closed expressions that run on the
local node. Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types and exceptions. Moreover, there are
primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, the three constructs 
\CONTRACT\ \TYPE\ \EXPR, \TOSTRING\  \EXPR, and \FROMSTRING\ \TYPE\
\EXPR\ are casts between untyped and typed representations of
contracts and their parameters.

@@

\begin{figure}[tp]
  \begin{align*}
    \LET\ \OPH &=  \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
         & \MATCH\ \QRY\ \GETSTATUS\OPH\ \WITH \\
         & \mid \STATUSPENDING \to \textrm{not yet included, continue waiting} \\
         & \mid \STATUSINCLUDING (n) \to \textrm{included with $n$
           blocks confirmation}\\
         & \mid \STATUSTIMEOUT \to \textrm{lost operation, try again}
  \end{align*}
  \caption{Pseudocode example for a token transfer}
  \label{fig:pseudocode-example}
\end{figure}
Figure~\ref{fig:pseudocode-example} shows an expression that invokes a
contract at \PUH\ with argument \STRING.

Subtyping in Figure~\ref{fig:subtyping} etablishes that public hashes
and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. Subtyping for function types has the usual co-/contravariant definition.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\TPUH \SubType \TADDR}

@@

    \inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}

    \inferrule{\TYPEU' \SubType \TYPE' \\ \TYPE \SubType \TYPEU}{
      \TYPE' \to \TYPE \SubType \TYPEU' \to \TYPEU}
  \end{mathpar}
  \caption{Subtyping}
  \label{fig:subtyping}
\end{figure}
Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

@@

    \inferrule{\JTypeExpr\TEnv\EXPR\TADDR}{\JTypeExpr\TEnv{\CONTRACT\
        \TYPE\ \EXPR}{\TOPTION (\TCONTRACT\ \TYPE)}}

    \inferrule{}{\JTypeExpr\TEnv\NTEZ\TTEZ}

@@

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \JTypeExpr\TEnv{\EXPR'}{\TEXCEPTION\to\TYPE} }{
      \JTypeExpr\TEnv{\TRY\ \EXPR\ \EXCEPT\ \EXPR'}\TYPE }
  \end{mathpar}
  \caption{Typing rules for expressions}
  \label{fig:typing-expressions}
\end{figure}
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TPUK \\
      % \JTypeExpr\TEnv{\EXPR_3}{\TCONTRACT\ \TUNIT} \\
      \JTypeExpr\TEnv{\EXPR_4}\TUNIT \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
    }

@@

\begin{definition}%[State of a node]
  The \emph{state of a node} is a tuple
  $\NODE = [ \EXPRS, \ACCOUNTS
  % , \OPERATIONS
  ]$, where $\EXPRS$ is a
  set of programs and $\ACCOUNTS \subseteq \TPAK \times \TPUK$  is a set of
  implicit accounts.
  % , and
  % $\OPERATIONS \subseteq \ECS \times \TOP $ is a set of pairs of
  % expression contexts and operations.
\end{definition}

>>>>

We write $\emptyset$ for the empty set and $\mathbf{e :: s}$ to
decompose a set into an element $\mathbf{e}$ and a set
$\mathbf{s}$. We generally use lowercase boldface for metavariables
ranging over values of a certain syntactic category, e.g., \PUK\ for
public keys, and the corresponding uppercase name for the
corresponding type or the set of these values (as in \TPUK).

@@

The local node offers operations to transfer tokens from one account
to another, to invoke a contract, and to originate a contract on the blockchain.
\begin{align*}
  \OP &::= \TRANSFER[\PARAMETER]\NTEZ\PUK\ADDR\MTEZ
  \\&\mid \ORIGINATE\NTEZ\PUK\CODE\INIT\MTEZ
\end{align*}
In the transfer, \NTEZ\ is the amount of tokens transferred, \PUK\ is the public key of the
sender, \ADDR\ is either a public 
key for an implicit account (in case of a simple transfer)  or a
public hash for a smart contract (for an invocation), \PARAMETER\ is
the argument passed to the smart contract, 
which is empty for a simple transfer, and \MTEZ\ is
the amount of tokens for the transaction fee. In originate,  \CODE\ is
the script of a smart contract and \INIT\ is the initial value of the
contract's storage. Each operation returns an \emph{operation hash}
\OPH, on which we can query the status of the operation.

The local node offers several ways to query the current state of
the blockchain.
\emph{Queries} are defined by the following grammar:
\begin{align*}
  \QRY & ::=
  % \GETCOUNTER\PUK  
  %        \mid
         \GETBALANCE\ADDR
  \mid \GETSTATUS\OPH
  \mid \GETSTORAGE\PUH
  % \mid \GETCODE\PUH 
  % \mid \GETTYPE\PUH
  \mid \GETCONTRACT\OPH
\end{align*}
We obtain the balance associated with an implicit account or a
contract by its public key or public hash, respectively; the status of
a submitted operation by its operation hash; the stored value of a
contract by its public hash; and the code of a contract by the
operation hash of its originating transaction.

\begin{definition}%[State of a node]
  The \emph{state of a node} is a pair
  $\NODE = [ \EXPRS, \ACCOUNTS
  % , \OPERATIONS
  ]$, where $\EXPRS$ is a
  set of programs and $\ACCOUNTS \subseteq \TPAK \times \TPUK$  is a set of
  implicit accounts.
  % , and
  % $\OPERATIONS \subseteq \ECS \times \TOP $ is a set of pairs of
  % expression contexts and operations.
\end{definition}

\begin{figure}[tp]
\begin{align*}
  \CONSTANT & ::= \INT \mid \FIX
              \mid \OPH \mid \PUH \mid\PUK \mid \CODE \mid
              \NTEZ \mid \SUNIT \mid \FALSE \mid \TRUE \\
  \STATUS & ::= \STATUSPENDING \mid \STATUSINCLUDING (\INT) \mid
            \STATUSTIMEOUT \\
  \ERROR &::= \ERRPRG \mid \ERRBAL \mid \ERRCOUNT \mid \ERRFEE \mid
           \ERRPUK \mid \ERRPUH \mid \ERRARG \mid \ERRINIT\\
	\EXPR & :: =  \CONSTANT \mid \STATUS \mid \ERROR \mid \VARIABLE \mid \lambda \VARIABLE. \EXPR\ 
	\mid \EXPR\EXPR  %\mid \FST\ \EXPR \mid \SND\ \EXPR 	%\mid \NI \EI
  \mid \EXPR\ \PLUS\ \EXPR\ \mid
  % \EXPR\ \MINUS\ \EXPR\ \mid
  \EXPR\ \EQUAL\ \EXPR\
  % \mid  \EXPR\ \LESS\ \EXPR
  \mid \EXPR\ \AND\ \EXPR\ \mid \EXPR\ \OR\ \EXPR\ \mid \NOT\ \EXPR
  \\& \mid (\EXPR,\EXPR)\mid \NIL\ \mid \CONS\ \EXPR\ \EXPR \mid \LEFT\ \EXPR \mid\RIGHT\
  \EXPR \mid \SOME\ \EXPR \mid \NONE
  \mid \MATCH\ \EXPR\ \WITH\ \PATTERN\to\EXPR \dots 
  \\& \mid \RAISE\ \EXPR \mid \TRY\ \EXPR\ \EXCEPT\ \EXPR \mid \CAST\EXPR\TYPE\TYPEU
  \\& \mid \QOP\  \EXPR \mid \TRANSFER[\EXPR]\EXPR\EXPR\EXPR\EXPR
  \\& \mid \ORIGINATE\EXPR\EXPR\EXPR\EXPR\EXPR
  % \\&\mid \CONTRACT\ \TYPE\ \EXPR
  \\& \mid \TOSTRING\ \EXPR \mid \FROMSTRING\ \TYPE\ \EXPR
  \\
  \PATTERN &::= \VARIABLE \mid (\PATTERN, \PATTERN) \mid \NIL \mid \CONS\ \PATTERN\ \PATTERN \mid \LEFT\ \PATTERN \mid\RIGHT\
             \PATTERN \mid \SOME\ \PATTERN \mid \NONE \\
  &\mid \FALSE \mid \TRUE \mid \STATUS \mid \ERROR\\
  \QOP & ::= \GETBALANCE{} \mid \GETSTATUS{} \mid \GETSTORAGE{} \mid \GETCODE{} \mid \GETCOUNTER{}
         \mid \GETTYPE{}
  \\[2ex]
  \TYPE, \TYPEU & ::=
                  \TPUH \mid
                  \TPUK \mid
                  \TADDR \mid 
                  \TCONTRACT\ \TYPE \mid
                  \TCODE\ \TYPE\ \TYPEU \mid \TTEZ \mid
                  \TOPH \mid
                  \TSTATUS \mid \TEXCEPTION \\
  & \mid \TINT \mid \TUNIT \mid \TBOOL \mid \TSTRING \mid \TYPE\to\TYPEU \mid \TPAIR\ \TYPE\ \TYPEU \mid \TLIST\ \TYPE
    \mid \TSUM\ \TYPE\ \TYPEU \mid \TOPTION\ \TYPE 
\end{align*}
  \caption{Syntax of expressions, \EXPR, and types, \TYPE}
  \label{fig:syntax-expressions}
\end{figure}
All queries and operations are started by programs, i.e., closed expressions that run on the
local node. Figure~\ref{fig:syntax-expressions} defines the syntax of
expressions, which describe the usual constructs of call-by-value PCF
with sum, product, list, and option types, exceptions and fixpoint. Moreover, there are
primitive types and constants \CONSTANT\ to support blockchain
operations, as well as several exceptional values collected in \ERROR.
Pattern matching is the only means to decompose values, cf.\ \PATTERN.
There is syntax to initiate transfers and to originate contracts as
well as for the queries. Finally, there is a type cast
$\CAST\EXPR\TYPE\TYPEU$, which is only applicable to pairs of
types related in the reflexive, transitive, and symmetric closure of
the subtyping relation in Figure~\ref{fig:subtyping}. Upcasts will
just go through, but downcasts may raise an exception, as usual.

@@

\begin{figure}[tp]
  \begin{align*}
    \LET\ \mathit{oph} &=  \TRANSFER\NTEZ\PUK{\PUK'}\MTEZ \\
         & \FIX \lambda f.\lambda x. \MATCH\ \GETSTATUS{\mathit{oph}}\ \WITH \\
         & \mid \STATUSPENDING \to \text{Time.sleep(1)}; f () \textrm{\qquad(* not yet included,
           continue waiting *)} \\
         & \mid \STATUSINCLUDING (n) \to \text{if } n < 50\text{ then
           Time.sleep(1)};f ()\textrm{\quad(* included with $n$
           blocks confirmation *)}\\
         & \mid \STATUSTIMEOUT \to \textrm{(* lost operation, try
           again *)}
  \end{align*}
  \caption{Pseudocode example for a token transfer}
  \label{fig:pseudocode-example}
\end{figure}
Figure~\ref{fig:pseudocode-example} shows an incomplete example
expression that transfers tokens from $\PUK$ to $\PUK'$ and then
checks the status of the operation. The outer loop to raise the fee
and retry is omitted.

Our type system does not include subtyping. The relation in
Figure~\ref{fig:subtyping} just etablishes pairs of types that can be
subjected to casting. A cast from $\TYPE$ to $\TYPEU$ is only allowed
if $\TYPE \SubType^* \TYPEU$ (upcast) or $\TYPEU \SubType^* \TYPE$
(downcast). Upcasts always succeed, but downcasts may fail at run time.
In particular,  public hashes and public keys can both stand for addresses. Moreover, a smart
contract with parameter type $\TYPE$ is represented by its public hash
at run time. The corresponding downcast must check whether the public
hash is valid and has the expected parameter type.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\TPUH \SubType \TADDR}

@@

    \inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}
  \end{mathpar}
  \caption{Subtyping}
  \label{fig:subtyping}
\end{figure}
Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

@@

    % \inferrule{\JTypeExpr\TEnv\EXPR\TADDR}{\JTypeExpr\TEnv{\CONTRACT\
    %     \TYPE\ \EXPR}{\TOPTION (\TCONTRACT\ \TYPE)}}
    %     
    \inferrule{}{\JTypeExpr\TEnv\NTEZ\TTEZ}

@@

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \JTypeExpr\TEnv{\EXPR'}{\TEXCEPTION\to\TYPE} }{
      \JTypeExpr\TEnv{\TRY\ \EXPR\ \EXCEPT\ \EXPR'}\TYPE }

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \TYPE \SubType \TYPEU \vee \TYPEU \SubType \TYPE
    }{
      \JTypeExpr\TEnv{\CAST\EXPR\TYPE\TYPEU}\TYPEU
    }
  \end{mathpar}
  \caption{Typing rules for expressions}
  \label{fig:typing-expressions}
\end{figure}
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TPUK \\
      % \JTypeExpr\TEnv{\EXPR_3}{\TCONTRACT\ \TUNIT} \\
      \JTypeExpr\TEnv{\EXPR_4}\TUNIT \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
    }

@@
